---
title: "Data 607- Final Project"
author: "Meaghan Burke"
date: "May 12, 2018"
output: html_document
---
### **Data 607- Final Project:**

#### **Introduction:**
I currently work as a business intelligence analyst supporting both sales and product managers for a fin-tech firm. Much of my job is using data to help the two groups make informed product and client decisions. For this final project, I would will apply the skills learned in this class to a sampling of data from work to see if I can find actionable insights.

#### **Project Goals- RFM:**

-  Segment the customers based on a RFM analysis. Currently clients are mapped based on industry and division. 
-  RFM (recency, frequency, monetary) analysis is used to determine a customer ranking by examining how recently a customer has purchased (recency), how often they purchase (frequency), and how much the customer spends (monetary). 
-  I plan to segment the data into actionable subsets (asset classes) and perform the RFM "peer analysis

#### **Data Source(s):**

-  Sales data aggregates
-  Client metadata aggregates


#### **Overview:**

-  Collect account usage data, available in monthly aggregates
-  Combined all csv files with linux "cat" command
-  Collect metadata information from the warehouse
-  Bring the aggregated data into R Studio
-  Anonymize the sensitive client information
-  Clean the data (very messy) with dplyr and tidyr
-  Perform an exploratory analysis on the data
-  Perform a RFM analysis on the data
-  Graphically represent my findings using GGPlot2
-  Attempt to answer to the "Goals" listed above

```{r setup, include=FALSE}
library(dplyr)
library(tidyr)
library(knitr)
library(ggplot2)
library(stringr)
```


#### **Obtain The Data:**

- 2017 aggregated sales data (aggregated offline with the linux "cat" command)
- Category mapping file "ncm"
- Asset class mapping file "dl.ac"

```{r echo = TRUE}
raw.data  <- '//corp.bloomberg.com/ny-dfs/Global Data/Other Shared Folders/Worldwide/Global Data/CRA/BI Data Sets/DLME/DLPS/CSV_Meaghan_Versions/csv/2017 Data/combined_2017.csv'

ncm <- "//corp.bloomberg.com/ny-dfs/Global Data/Other Shared Folders/Worldwide/Global Data/CRA/Work Area/Personal Work Area/Meaghan/Tech Working Group/Sales_Prediction/ncm_mapping.csv"

dl.ac <- "//corp.bloomberg.com/ny-dfs/Global Data/Other Shared Folders/Worldwide/Global Data/CRA/Work Area/Personal Work Area/Meaghan/Tech Working Group/Sales_Prediction/dl_asset_class_mapping.csv"

```

```{r echo = TRUE, warning = FALSE, message = FALSE}

initial.data <- read.table(raw.data,  sep=',', encoding='latin-1',header = FALSE, fill = TRUE)

ncm.mapping  <-read.table(ncm,sep=',', encoding='latin-1')%>%
                rename(Category =V1, Mapping = V2)%>%
                mutate_if(is.factor, as.character)

dl.ac.mapping = read.table(dl.ac,   sep=',', encoding='latin-1')%>%
                rename(Asset.Class = V1, Ac.Mapping = V2)%>%
                mutate_if(is.factor, as.character)


```
#### **Raw data summary:**
-  The raw dataset is large at approximately 1.4 million rows and 18 columns
-  Examine the data types of each column to determine if transformations are needed
```{r echo = TRUE}
dim(initial.data)

sapply(initial.data, class)
```

#### **Scrub The Data:**
-  rename the columns 
-  Remove unneeded columns
-  Cast the number columns to numerics
-  filter for the relevant subset of data 
```{r echo = TRUE, warning = FALSE, message = FALSE}
clean.data <- initial.data%>%
                rename(Date = V1, Firm.Num = V2, Division = V4,
                       Account.Num = V6, Asset.Class = V12, Category = V11,Billing.Model = V9,
                       Hit.Type = V14, Billed.Qty = V15, Service = V10) %>%
                select(-c(V3, V5, V7, V8,V13,V16,V17,V18))%>%
                mutate_at(c(2, 4,10), as.numeric) %>%
                mutate_if(is.factor, as.character)%>%
                filter(Billed.Qty > 1,
                       Date != "Date",
                       Billing.Model == "NCM")%>%
                select(-c(Billing.Model))%>%
                filter(Hit.Type == "Unique"|
                       Hit.Type == "Uniques")%>%
                mutate(Division = ifelse(Division == "Sell Side"| Division == "Sellside", 2,
                                  ifelse(Division == "Buy Side" | Division == "Buyside", 1,
                                  ifelse(Division == "Custody"| Division == "Custodial", 3,
                                  ifelse(Division == "Other",5,0))))) %>%
                mutate(Date = ifelse(Date == "20171231", 12,
                              ifelse(Date == "20171130", 11,
                              ifelse(Date == "20171031.0", 10,
                              ifelse(Date == "20170930.0", 9,
                              ifelse(Date ==  "20170831", 8,
                              ifelse(Date == "20170731.0", 7,0)))))))%>%
                mutate(Service = ifelse(Service == 'Ad Hoc',1,0))%>%
                filter(Date !=0)%>%
                na.omit() %>%
                mutate(Firm.Num = round(Firm.Num/10),
                         Account.Num =round( Account.Num/10),
                       Billed.Qty = round(Billed.Qty/10))%>%
                left_join(ncm.mapping)%>%
                left_join(dl.ac.mapping)%>%
                select(-c(Category,Asset.Class,Hit.Type))%>%
                rename(Category = Mapping, Asset.Class = Ac.Mapping)%>%
                arrange(desc(Billed.Qty))
kable(head(clean.data,15),  caption = "Cleaned Data")
```
#### **Summary of the "scrubbed" data:**
- The cleaned data subset has ~421 K rows and 8 columns 
- A sumary of the cleaned data is below
  -  the quantity has a wide distribution
- The bubble chart and bar chart below show the mean quantity by asset class and category  
  -  Asset class 1, 2 & 3 have the highest usage 
  -  Category 8,7 & 9 seem to be the most used per asset class
```{r echo= TRUE}
dim(clean.data)
summary(clean.data)
```
```{r echo = TRUE}
group.df <- clean.data %>%
            select(-c(Firm.Num, Division, Service))%>%
            group_by(Asset.Class, Category)%>%
            na.omit() %>%
            summarise(Billed.Qty = mean(Billed.Qty))%>%
            arrange(desc(Billed.Qty))
ggplot(group.df, aes(x = Asset.Class, y = Category, label = Category)) +
    geom_point(aes(size = Billed.Qty, colour = Asset.Class)) + 
    geom_text(hjust = 1, size = 2) +
    scale_size(range = c(1,15)) +
    theme_bw()
```

```{r echo = TRUE}
g <- ggplot(group.df, aes(reorder(Asset.Class, -Billed.Qty)))
g + geom_bar(aes(fill= Category), width = 0.5) +
  theme(axis.text.x = element_text(angle=65, vjust=0.6)) +
  labs(x = 'Asset.Class',
       y = 'Qty',
       title="Asset Class by Category Bar Chart", 
       caption="Source: Cleaned Dataset")
```

#### **Explore the Data:**

-  Repivot the data from long to wide. The 6 months of data from rows to columns
-  Calculate the weighted average usage over a 6 month period 
-  Calculate the Frequency over the 6 month period
-  Calculate the Recency over the 6 month period
-  Group by asset class
-  Summarize by RFM metrics: 
    - Frequency(Max),the more frequent the better
    - Recency (Max), the more recent the better
    - Amount(Sum), the higher the better

```{r echo = TRUE, warning = FALSE, message=FALSE}
clean.data.w <- reshape(clean.data, direction = "wide", idvar = c("Firm.Num", "Division","Account.Num", "Service","Category","Asset.Class"), timevar = "Date")
kable(head(clean.data.w,15),  caption = "Wide Usage Table")
```

```{r echo = TRUE}
rfm.table <- clean.data.w %>%
                rename('7' = Billed.Qty.7,
                       '8' = Billed.Qty.8,
                      '9' = Billed.Qty.9,
                       '10' =Billed.Qty.10,
                       '11' =Billed.Qty.11,
                       '12' =Billed.Qty.12)%>%
                replace(is.na(.),0)%>%
                mutate(total = rowSums(.[7:12]))%>%
                mutate(Wgt.Avg = round(((rowSums(.[7])/rowSums(.[7:12])*rowSums(.[7]))+
                                 (rowSums(.[8])/rowSums(.[7:12])*rowSums(.[8]))+
                                 (rowSums(.[9])/rowSums(.[7:12])*rowSums(.[9]))+
                                 (rowSums(.[10])/rowSums(.[7:12])*rowSums(.[10]))+
                                 (rowSums(.[11])/rowSums(.[7:12])*rowSums(.[11]))+
                                 (rowSums(.[12])/rowSums(.[7:12])*rowSums(.[12])))))%>%
                mutate(Last.Usage = ifelse(rowSums(.[12]) !=0,5,
                                    ifelse(rowSums(.[11]) !=0,4,      
                                    ifelse(rowSums(.[10]) !=0,3,  
                                    ifelse(rowSums(.[9]) !=0,2,  
                                    ifelse(rowSums(.[8]) !=0,1,  
                                    ifelse(rowSums(.[7]) !=0,0,0))))))) %>%
                mutate(Frequency = rowSums(.[,7:12] >0))%>%
                select(c(Firm.Num, Division, Account.Num, Service, Category, Asset.Class, total, Wgt.Avg, Last.Usage, Frequency))%>%
                group_by(Firm.Num, Division, Account.Num, Service, Asset.Class)%>%
                summarize(Frequency = max(Frequency),
                          Recency = max(Last.Usage),
                          Amount = sum(Wgt.Avg))%>%
              arrange(desc(Amount))

kable(head(rfm.table,15),  caption = "Restructured Usage Table")


```


```{r}
ggplot(rfm.table, aes(x=reorder(Asset.Class, -Amount), y= Amount)) + 
  geom_bar(stat="identity", width=.5, fill="slategray2") +
  geom_text(aes(label=Amount))+
  labs(x='Asset Class', 
       y='Weighted Qty',
       title="Weighted Usage Qty by Asset Class", 
       caption="Source: RFM Table Dataset") + 
  theme(axis.text.x = element_text(angle=65, vjust=0.6))
```

#### **Model:**
-  Loop throgh the dataframe by asset class to get the relevant RFM Scores (peer group analysis)
    - Ntile() was used to partition the data. Great function!
-  **The higher the score the better the customer **
  -  Bougth most recently, with frequency and a lot
```{r echo = TRUE}
temp.list = list()
df.total = data.frame()
unique <- unique(unlist(rfm.table$Asset.Class))

for (i in 1:length(unique)){
  temp.df <- subset(rfm.table, Asset.Class == unique[i])
  temp.df$Amt.tile <- ntile(temp.df$Amount, 10)
  temp.df$Frq.tile <- ntile(temp.df$Frequency, 4)
  temp.df$Rec.tile <- ntile(temp.df$Recency, 4)
  
  temp.list[[i]]<- temp.df
  
}
rfm.final <- do.call(rbind, temp.list)%>%
            mutate_if(is.integer, as.character)%>%
            mutate(Score = paste(Amt.tile, Frq.tile, Rec.tile))%>%
            mutate(Score= str_replace_all(Score, fixed(" "), ""))%>%
            
            mutate(Score = as.integer(Score),
                   Amount = as.integer(Amount),
                   Recency = as.integer(Recency),
                   Frequency = as.integer(Frequency))%>%
            arrange(desc(Score))%>%
            mutate(Counter = 1)


kable(head(rfm.final,15),  caption = "RFM Table")
```
```{r echo = TRUE}
agg.df <- rfm.final%>%
          group_by(Asset.Class, Score)%>%
          summarise(Counter = sum(Counter))%>%
          arrange(desc(Counter))
kable(agg.df,  caption = "RFM Table By Score")
```


```{r echo = TRUE}
ggplot(agg.df, aes(x = Score, y = Asset.Class, label = Asset.Class)) +
    geom_point(aes(size = Counter, colour = Asset.Class)) + 
    geom_text(hjust = 1, size = 2) +
    scale_size(range = c(1,15)) +
    theme_bw()
```

#### **Conclusions- Customer Segmentation:**

https://www.putler.com/rfm-analysis/

-  Now that the customers have been scored, they can be segmented into different groups based on usage trends: 
  - Champions: Bought recently, buy often and spend the most!	
  - Loyal Customers:Spend good money with us often
  - Potential Loyalist:	Recent customers, but spent a good amount and bought more than once
  - Recent Customers:	Bought most recently, but not often.
  - Customers Needing Attention:	Above average recency, frequency and monetary values
  - About To Sleep:	Below average recency, frequency and monetary values
  - Can't Lose Them:	Made biggest purchases, and often. But haven't returned for a long time
  - Hibernating:	Last purchase was long back, low spenders and low number of orders


#### **Campions**
-  Bought recently, buy often and spend the most

```{r echo = TRUE}
campion <- rfm.final%>%
          filter(Amt.tile == 10,
                 Frq.tile == 4,
                 Rec.tile == 4)%>%
          arrange(desc(Score))
kable(head(campion, 20), caption = "Campion Customers")
```

#### **Customers Needing Attention:**
-  Above average recency, frequency and monetary values. May not have bought very recently though.

```{r echo = TRUE}
campion <- rfm.final%>%
          filter(Amt.tile == 10,
                 Frq.tile == 4,
                 Rec.tile == 2)%>%
          arrange(desc(Score))
kable(head(campion, 20), caption = "At Risk Customers")
```


